\begin{exercises}

\problem Let $A=\{1,2,3,4\}$ and let $B=\{a,b,c\}$.
Find the sets $A\times B$ and $B\times A$.

\problem Let $A$ be the set $\{a,b,c,d\}$.  Let $f$ be the
function from $A$ to $A$ given by the set of ordered pairs
$\{(a,b),(b,b),(c,a),(d,c)\}$, and let $g$ be the function
given by the set of ordered pairs $\{(a,b),(b,c),(c,d),(d,d)\}$.
Find the set of ordered pairs for the composition $g\circ f$.

\problem Let $A=\{a,b,c\}$ and let $B=\{0,1\}$.  Find all
possible functions from $A$ to $B$.  Give each function as
a set of ordered pairs.  (Hint: Every such function corresponds
to one of the subsets of $A$.)

\problem Consider the functions from $\Z$ to $\Z$ which are
defined by the following formulas.  Decide whether each
function is onto and whether it is one-to-one; prove your answers.  
\pparts{
   f(n)=2n&       g(n)=n+1&       h(n)=n^2+n+1\cr\noalign{\smallskip}
   s(n)=\left\{\begin{tabular}{ll}
            n/2,&\text{if $n$ is even}\cr
            (n+1)/2,&\text{if $n$ is odd}
          \end{tabular}
        \right.\hidewidth\cr
}

\problem Prove that composition of functions is an associative
operation.  That is, prove that for functions
$f\colon A\to B$, $g\colon B\to C$, and $h\colon C\to D$,
the compositions $(h\circ g)\circ f$ and $h\circ(g\circ f)$
are equal.

\problem 
Suppose that $f\colon A\to B$ and $g\colon B\to C$ are
functions and that $g\circ f$ is one-to-one.  
\ppart Prove that $f$ is one-to-one. (Hint: use a proof by contradiction.)
\ppart Find
a specific example that shows that $g$ is not necessarily
one-to-one.

\problem Suppose that $f\colon A\to B$ and $g\colon B\to C$,
and suppose that the composition $g\circ f$ is an onto
function.  
\ppart Prove that $g$ is an onto function.  
\ppart Find
a specific example that shows that $f$ is not necessarily
onto.



\end{exercises}



