
\begin{exercises}
\problem Express the set of prime numbers in set-builder notation.

\problem If we don't make the assumption that $a$, $b$, and $c$ are distinct,
then the set denoted by $\{a,b,c\}$ might actually contain either 1, 2, or 3 
elements.  How many different elements might the set
$\{\,a,\,b,\,\{a\},\,\{a,c\},\,\{a,b,c\}\,\}$
contain?  Explain your answer.

\problem Compute $A\cup B$, $A\cap B$, and $A\SETDIFF B$ for each of the following
pairs of sets
\pparts{
     A =\{a,b,c\},\ B=\emptyset \cr
     A =\{1,2,3,4,5\},\ B=\{2,4,6,8,10\} \cr
     A =\{a,b\},\ B=\{a,b,c,d\} \cr
     A =\{a,b,\{a,b\}\},\ B=\{\{a\},\{a,b\}\}
}

\problem Recall that $\N$ represents the set of natural numbers.  That is,
$\N=\{0,1,2,3,\dots\}$.  Let $X=\{n\in\N\st n\ge 5\}$,
let $Y=\{n\in\N\st n\le10\}$, and let $Z=\{n\in\N\st \text{$n$ is an even number}\}$.
Find each of the following sets:
\pparts{
     X\cap Y  & X\cup Y   & X\SETDIFF Y &\N\SETDIFF Z\cr
     X\cap Z  & Y\cap Z   & Y\cup Z   & Z\SETDIFF\N\cr
}


\problem Find $\POW\big(\{1,2,3\}\big)$. (It has eight elements.)

\problem Assume that $a$ and $b$ are elements and that $a\not=b$. Let $A$ and $B$ be the sets defined by
$A=\{\,a,\,\{b\},\,\{a,b\}\,\}$ and $B=\{\,a,\,b,\,\{a,\{b\}\}\,\}$.
Determine whether each of the following statements is true or false.  Explain
your answers.
\pparts{
    b\in A & \{a,b\}\SUB A & \{a,b\}\SUB B\cr 
    \{a,b\}\in B & \{a,\{b\}\}\in A & \{a,\{b\}\}\in B
}

\problem Since $\POW(A)$ is a set, it is possible to form the set
$\POW\big(\POW(A)\big)$.  What is $\POW\big(\POW(\emptyset)\big)\,$?
What is $\POW\big(\POW(\{a,b\})\big)\,$? (It has sixteen elements.)

\problem In the English sentence, ``She likes men who are tall, dark, and
handsome,'' does she like an intersection or a union of sets of men?
How about in the sentence, ``She likes men who are tall, men who are dark,
and men who are handsome?''

\problem If $A$ is any set, what can you say about $A\cup A\,$?
About $A\cap A\,$?  About $A\SETDIFF A\,$?  Why?

\problem Suppose that $A$ and $B$ are sets such that $A\SUB B$.
What can you say about $A\cup B\,$?  About $A\cap B\,$?
About $A\SETDIFF B\,$?  Why?

\problem Suppose that $A$, $B$, and $C$ are sets.  Show that
$C\SUB A\cap B$ if and only if $(C\SUB A)\AND (C\SUB B)$.

\problem\label{E-subtrans} Suppose that $A$, $B$, and $C$ are sets, and that
$A\SUB B$ and $B\SUB C$.  Show that $A\SUB C$.

\problem Suppose that $A$ and $B$ are sets such that $A\SUB B$.
Is it necessarily true that $\POW(A)\SUB \POW(B)\,$?
Why or why not?


\problem Let $M$ be any natural number, and let $P(n)$ be a
predicate whose domain includes all natural numbers
greater than or equal to $M$.  Suppose that $P(M)$ is true,
and suppose that $P(k)\IMP P(k+1)$ for all $k\ge M$.
Show that $P(n)$ is true for all $n\ge M$.


\end{exercises}






